---
title: "ulbphd_final"
author: "javad kouhestani"
date: "2025-12-29"
output:
  html_document: default
  pdf_document: default
---

## libraries

```{r}
library(readxl)
library(tidyverse)
library(psych)
library(lavaan)
library(ggplot2)
library(mediation)
library(interactions)
library(apaTables)
```

## Data wrangling

note: some of the decisions are for better readability of the codes in the markdown version. this can facilitate the communication with research partners.

```{r}
#reading the excel file
d_raw <- read_excel("Copy of Data PhD critical review paper.xlsx")
d <- d_raw
#making sure about the cleaning of the data using R's regular expressions (regex)
d <- d %>% mutate(across(matches("^antw(1[2-9]|[2-9][0-9])$"), ~ ifelse(.x %in% 1:7, .x, NA)))
#renaming the columns for better readability of the codes and avoiding mistakes
d <- d %>% rename(
    gender = antw2,
    nationality = antw3,
    language = antw4,
    ethnicity = antw5,
    religion = antw6,
    education = antw7,
    income = antw8,
    pol_placement = antw9,

    rwa01 = antw12,
    rwa02 = antw13,
    rwa03 = antw14,

    sdo01 = antw15,
    sdo02 = antw16,
    sdo03 = antw17,

    rwatown   = antw19,
    rwawork   = antw20,
    rwafamily = antw21,
    rwapeers  = antw22,
    
    check     = antw23,

    sdotown   = antw24,
    sdowork   = antw25,
    sdofamily = antw26,
    sdopeers  = antw27,

    sharedrealityt = antw30,
    sharedrealityw = antw31,
    sharedrealityf = antw32,
    sharedrealityp = antw33,

    influencet = antw34,
    influencew = antw35,
    influencef = antw36,
    influencep = antw37,

    enactt = antw38,
    enactw = antw39,
    enactf = antw40,
    enactp = antw41,

    prototypt = antw42,
    prototypw = antw43,
    prototypf = antw44,
    prototypp = antw45,

    connectt = antw46,
    connectw = antw47,
    connectf = antw48,
    connectp = antw49,

    avoid01 = antw52,
    avoid02 = antw53,
    avoid03 = antw54,

    warmth01 = antw55,
    warmth02 = antw56,
    warmth03 = antw57
  )
head(d)
dim(d)

#remove check NA's
d <- d %>% filter(check == 5)
dim(d) #result: 12 observations have been removed
#removing instructions using regex
d <- d %>% dplyr::select(-dplyr::matches("^instr"))

#missing data summary
##missing per variable
d %>% summarise(across(everything(), ~ mean(is.na(.x)))) %>% # first turning the data into true-false values and then calculating the mean
  pivot_longer(everything(), names_to = "variables", values_to = "missing_prop") %>%
  arrange(desc(missing_prop))
##missing per respondents
d %>% mutate(n_missing = rowSums(is.na(across(everything()))), missing_prop = n_missing / ncol(d)) %>%
  dplyr::select(id, n_missing, missing_prop) %>%
  arrange(desc(missing_prop))
##remove participants with high NA's:
d <- d %>% filter(!id %in% c(226, 337, 440, 425, 625))
dim(d) #now we have 608 observations

#class modifications
##checking the class of the variables
str(d)
##then changing them to the correct class
d <- d %>% mutate(
    id = as.integer(id),
    # nominal variables
    gender = factor(gender, levels = c(1, 2, 3), labels = c("m", "f", "x")),
    nationality = factor(nationality, levels = c(1, 2), labels = c("US", "other")),
    language = factor(language, levels = c(1, 2, 3, 4), labels = c("english", "spanish", "chinese", "other")),
    ethnicity = factor(ethnicity, levels = c(1,2,3,4,5,6,7),
                       labels = c("white/european american","black/african american","native american",
                                  "asian american","latinx/hispanic american", "native hawaiian/pacific islander","other")),
    religion = factor(religion,    levels = c(1,2,3,4,5,6,7),
                         labels = c("Christian","Muslim","Jew","Mormon","Hindu","Buddhist","Atheist")),
    pol_placement = factor(pol_placement, levels = c(1,2,3,4),
                           labels = c("Democrat","Republican","Independent","Other"))
    )

#final check
str(d)
```

## moderation preparation

#### confirmatory factor analysis

it is used for showing the robustness of the constructs, before making the composite scores.

```{r}
# simultaneous cfa for dependent variables as they are theoretically related (also it allows to measure the fit indices because with three items, df was 0 --> no fit index)
cfadv <- '
  avoid =~ avoid01 + avoid02 + avoid03
  warmth =~ warmth01 + warmth02 + warmth03
'
fitcfadv <- cfa(cfadv, data = d, std.lv = TRUE)
summary(fitcfadv, fit.measures = TRUE, standardized = TRUE)
fitMeasures(fitcfadv, c("cfi","tli","rmsea","srmr")) # all indices show fitness, df=15

# simultaneous cfa for mediators
cfam <- '
  rwa_ind =~ rwa01 + rwa02 + rwa03
  sdo_ind =~ sdo01 + sdo02 + sdo03
'
fitcfam <- cfa(cfam, data = d, std.lv = TRUE)
summary(fitcfam, fit.measures = TRUE, standardized = TRUE)
fitMeasures(fitcfam, c("cfi","tli","rmsea","srmr")) # without any modificating correlation, rmsea =.111, others were fit
##modification indices
mi <- modindices(fitcfam)
mi_cfam <- mi[order(mi$mi, decreasing = TRUE), ]
head(mi_cfam, 20) 
```

#### reliability check with cronbach's alpha and making composite scores

```{r}
#defining the constructs
## individual-level constructs
rwa_items <- c("rwa01", "rwa02", "rwa03")
avoid_items <- c("avoid01", "avoid02", "avoid03")

## context(group)-level constructs
rwa_context_items <- c("rwafamily", "rwapeers")
enact_context_items <- c("enactf", "enactp")                

#cronbach's alpha for reliability check
alpha_rwa <- alpha(d[, rwa_items])
alpha_avoid <- alpha(d[, avoid_items])
alpha_rwa_ctx <- alpha(d[, rwa_context_items])
alpha_enact_ctx <- alpha(d[, enact_context_items])

# making composite scores
d$rwa_comp <- rowMeans(d[, rwa_items], na.rm = TRUE)
d$avoid_comp <- rowMeans(d[, avoid_items], na.rm = TRUE)
d$rwa_context_comp <- rowMeans(d[, rwa_context_items], na.rm = TRUE)
d$enact_context_comp <- rowMeans(d[, enact_context_items], na.rm = TRUE)


#table for summarizing the results
rel_check <- data.frame(
  Scale = c("RWA", "Avoidance", "RWA (context)", "Enact"),
  Alpha = c(
    alpha_rwa$total$raw_alpha,
    alpha_avoid$total$raw_alpha,
    alpha_rwa_ctx$total$raw_alpha,
    alpha_enact_ctx$total$raw_alpha
  ),
  Mean = c(
    mean(d$rwa_comp, na.rm = TRUE),
    mean(d$avoid_comp, na.rm = TRUE),
    mean(d$rwa_context_comp, na.rm = TRUE),
    mean(d$enact_context_comp, na.rm = TRUE)
  ),
  SD = c(
    sd(d$rwa_comp, na.rm = TRUE),
    sd(d$avoid_comp, na.rm = TRUE),
    sd(d$rwa_context_comp, na.rm = TRUE),
    sd(d$enact_context_comp, na.rm = TRUE)
  )
)
rel_check

#final result: all alphas are above 0.75
```

####control variables

```{r}
#checking the correlation with the outcomes
##for ordinal/numerical variables
num_cntrl <- c("income","education", "age")
cor(d[num_cntrl], d$avoid_comp, use = "pairwise.complete.obs") #all correlations are negligable

##gender
summary(lm(avoid_comp ~ gender, data = d)) #(b = −0.20, p = .011) small effect

##nationality
summary(lm(avoid_comp ~ nationality, data = d)) #not significant

##ethnicity
summary(lm(avoid_comp ~ ethnicity, data = d)) #not significant

##religion
summary(lm(avoid_comp ~ religion, data = d))  #not significant

##political placement
summary(lm(avoid_comp ~ pol_placement, data = d))  # republicans/democrats (b = -0.50, p < .001), small effect

#final results: only gender and political placement show the significance. however, i am not sure about considering political placement as a control variable as it can play the role of a predictor itself (i need to consult with the supervisor about that). regarding gender, i will add it to the main context level regression analysis, yet for having more parsimony, it will not be added as the control variable to all of the regression models. 


```

#### mean centering

```{r}
d$rwa_comp_c <- scale(d$rwa_comp, center = TRUE, scale = FALSE)

d$rwa_context_comp_c <- scale(d$rwa_context_comp, center = TRUE, scale = FALSE)
d$enact_context_comp_c <- scale(d$enact_context_comp, center = TRUE, scale = FALSE)

d$enactf_c <- scale(d$enactf, center = TRUE, scale = FALSE)
d$enactp_c <- scale(d$enactp, center = TRUE, scale = FALSE)

d$rwafamily_c <- scale(d$rwafamily, center = TRUE, scale = FALSE)
d$rwapeers_c <- scale(d$rwapeers, center = TRUE, scale = FALSE)
```

## linear regression

###### RWA group context -\> DV's

```{r}
# main
rwacntxt_avoid <- lm(avoid_comp ~ rwa_context_comp_c, data = d)
summary(rwacntxt_avoid)
## group level RWA is positively and significantly associated with out-group avoidance attitudes (r2 = 0.05, b = 0.16, p < .001).

#control
rwacntxt_avoid_cntrl <- lm(avoid_comp ~ rwa_context_comp_c + gender, data = d)
summary(rwacntxt_avoid_cntrl)
## same prediction, independence from gender (b and p), (women showed lower avoid)

```

## mediations preps

```{r}
#prep for mediation --> making these variables numeric
d$rwa_context_comp_c <- drop(d$rwa_context_comp_c)
d$enact_context_comp_c <- drop(d$enact_context_comp_c)
```

## mediation

```{r}
# rwa group context * enact context --> avoid
mrwa_enact_avoid <- lm(rwa_comp ~ rwa_context_comp_c, data = d)
summary(mrwa_enact_avoid)
yrwa_enact_avoid <- lm(avoid_comp ~ rwa_context_comp_c + rwa_comp, data = d)
summary(yrwa_enact_avoid)

med_rwaencavoid <- mediate(
  model.m = mrwa_enact_avoid,
  model.y = yrwa_enact_avoid,
  treat = "rwa_context_comp_c",
  mediator = "rwa_comp",
  boot = TRUE
)
summary(med_rwaencavoid)
```

## moderations + mediation

#### RWA group context \* enact -\> rwa med -\> DV's

using process model 8 (Hayes, 2013) i ran the process code in another r script (it is not a package in r). this can be done in SPSS easier as it is embedded there.

```{r, eval = FALSE}
# this can work only after running the PROCESS code (from this website: https://haskayne.ucalgary.ca/CCRAM/resource-hub). also this can be done on SPSS as the PROCESS is embedded in SPSS.
process(
  data = d,
  y = "avoid_comp",
  x = "rwa_context_comp_c",
  m = "rwa_comp",
  w = "enact_context_comp_c",
  model = 8,
  center = 2,
  moments = 1,
  modelbt = 1,
  boot = 10000,
  seed = 654321
)
```

###### plot (moderation)

```{r}
# rwa group context * enact context --> avoid
pprwa_enact_avoid <- lm(rwa_comp ~ rwa_context_comp_c * enact_context_comp_c, data = d)

###plot
interact_plot(
  model = pprwa_enact_avoid,
  pred = rwa_context_comp_c,
  modx = enact_context_comp_c,
)

```

## repeated-measures ANOVA

```{r}
# repeated-measures anova
d_longed <- d %>% 
  pivot_longer(cols = c(enactt, enactw, enactf, enactp),
  names_to = "domain",
  values_to = "compcen"
)
anova_cen <- aov(compcen ~ domain + Error(id/domain), data = d_longed)
summary(anova_cen)
# a one-way repeated-measures ANOVA shows a significant difference between the means, F(3, 2424) = 58.66, p < .001

# post-hoc pairwise comparisons
pairwise.t.test(
  d_longed$compcen,
  d_longed$domain,
  paired = TRUE,
  p.adjust.method = "holm"
)
# all are significant except for family and peers

# calculating means
d_longed %>% group_by(domain) %>%
  summarise(mean = mean(compcen, na.rm = TRUE), sd = sd(compcen, na.rm = TRUE))

# visualization
ggplot(d_longed, aes(x = domain, y = compcen)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = .15) +
  scale_x_discrete(labels = c(
    enactf = "Family",
    enactp = "Friends",
    enactt = "Town",
    enactw = "Colleagues"
  )) +
  labs(
    x = "Social domain",
    y = "norm enactment (mean)",
    title = "Mean norm enactment across social domains"
  ) +
  theme_minimal()

```

## correlation table of the variables (apa style)

```{r}
#using apatables package
apa.cor.table( d[, c("rwa_comp", "rwa_context_comp", "enact_context_comp", "avoid_comp")],
  filename = "Table1_Correlationsfinal2.doc"
)

```
